---
title: "Predicting the Best Times to Spot Wildlife"
subtitle: "A confidence-weighted scoring function for the ecotourism package"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: flatly
    self-contained: true
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## The question

The hard task asked for a function that takes occurrence and weather
data and predicts the top five days and times to spot an organism.

The first question is what *predict* means here. The package contains no
future weather forecasts. What it contains is ten years of sighting
records with timestamps, the weather conditions on those days, and a set
of top weather stations matched to each organism. The approach taken
here is:

> Find the month × hour combinations that have historically produced the
> most sightings, weight those combinations by how much weather
> conditions on sighting days differed from the baseline, and report
> each recommendation alongside an honest confidence score that reflects
> how much data backs it up.

This is the prediction — if these seasonal and weather conditions
repeat, this is when you are most likely to see the organism.

------------------------------------------------------------------------

## Setup

```{r}
#| label: setup

library(ecotourism)
library(dplyr)
library(tidyr)
library(lubridate)

data("glowworms")
data("gouldian_finch")
data("manta_rays")
data("orchids")
data("weather")
data("top_stations")
```

------------------------------------------------------------------------

## Understanding the data

Before writing the function, three things needed to be established from
the data.

### Record types and the midnight artifact

Inspecting `record_type` across all four organisms revealed an important
data quality issue:

```{r}
#| label: record-types

# Record types per organism
lapply(
  list(glowworms = glowworms, gouldian_finch = gouldian_finch,
       manta_rays = manta_rays, orchids = orchids),
  function(df) table(df$record_type)
)
```

Manta rays stand out immediately — 675 of 953 records are
`MACHINE_OBSERVATION`. The following confirms these are not real
behavioural sightings:

```{r}
#| label: machine-obs-evidence

manta_rays %>%
  group_by(record_type) %>%
  summarise(
    n           = n(),
    pct_hour_0  = round(mean(hour == 0) * 100, 1),
    median_hour = median(hour)
  )
```

**100% of `MACHINE_OBSERVATION` records are recorded at hour 0.** This
is not manta ray behaviour — it is the default timestamp written by
acoustic tags and satellite trackers when no exact time is available.
These records are excluded from all time-of-day analysis in the
function.

### The join strategy

The occurrence datasets each contain a `ws_id` column — the nearest
weather station to each sighting. Rather than joining directly on
`ws_id` and `date` (which produces few matches because occurrence
stations often differ from weather stations), the `top_stations` table
is used as a bridge. It maps each organism to its two or three most
relevant weather stations, giving much better weather coverage.

Where multiple stations cover the same date, weather variables are
averaged across stations before joining. This also resolves the
many-to-many join warning that arises when a single sighting date
matches multiple station records.

### Hour distributions and ecological plausibility

The hour distribution for each organism reveals its activity pattern:

```{r}
#| label: hour-distributions

cat("--- Glowworms ---\n");      table(glowworms$hour)
cat("\n--- Gouldian Finch ---\n"); table(gouldian_finch$hour)
cat("\n--- Manta Rays (human only) ---\n")
manta_rays %>%
  filter(record_type == "HUMAN_OBSERVATION") %>%
  pull(hour) %>% table()
cat("\n--- Orchids ---\n");        table(orchids$hour)
```

Rather than hardcoding an activity window per organism, the function
derives the plausible hour range directly from the data — the central
80% of sightings (10th to 90th percentile of the hour distribution). A
cosine taper is applied outside this window so there is no hard cliff at
the boundary.

### Data sparsity

How sparse is the month × hour grid for each organism?

```{r}
#| label: sparsity

organisms <- list(
  glowworms      = glowworms,
  gouldian_finch = gouldian_finch,
  manta_rays     = manta_rays,
  orchids        = orchids
)

for (name in names(organisms)) {
  cat("\n---", name, "---\n")
  organisms[[name]] %>%
    filter(hour != 0, record_type != "MACHINE_OBSERVATION") %>%
    count(month, hour) %>%
    summarise(
      total_cells      = n(),
      median_n         = median(n),
      cells_with_1     = sum(n == 1),
      cells_with_5plus = sum(n >= 5)
    ) %>% print()
}
```

Glowworms and manta rays are very sparse — most month × hour cells
contain only a single sighting. Orchids and gouldian finch are dense. A
naive frequency ranking treats a cell with 1 sighting the same as one
with 50. The confidence score addresses this directly.

------------------------------------------------------------------------

## The function

The function combines three components into a final score for every
month × hour combination:

**Composite score** — weighted average of month frequency (40%), hour
frequency (40%), and weather favourability (20%). Weather weight is
self-calibrating: variables are weighted by Cohen's d, the standardised
difference between sighting-day conditions and baseline. If weather
barely differs on sighting days, it contributes almost nothing to the
score.

**Confidence score** — exponential saturation curve `1 - exp(-n / k)`,
where `n` is the number of sightings in that month × hour cell and
`k = 5` by default. This is borrowed from reliability engineering. It is
honest about sparse data: 1 sighting scores 0.18, 5 sightings scores
0.63, 10 scores 0.86, 20+ approaches 0.98 but never reaches 1.0.

**Ecological plausibility score** — derived from the data, not
hardcoded. The central 80% of the hour distribution defines the
plausible window. Hours inside score 1.0; hours outside decay smoothly
via a cosine taper to a floor of 0.2.

**Final score** = `composite × (0.6 × confidence + 0.4 × plausibility)`

This means:

-   High confidence + plausible hour → high score, recommended strongly
-   Low confidence + plausible hour → moderate score, flagged as sparse
    but biologically consistent
-   High confidence + implausible hour → penalised, flagged as caution
-   Low confidence + implausible hour → penalised hard

```{r}
#| label: function-definition

predict_best_spotting_times <- function(occurrence_data,
                                        weather_data,
                                        top_stations,
                                        organism_name,
                                        n      = 5,
                                        conf_k = 5) {

  # ── 1. Remove machine observations ────────────────────────────────────────
  # Evidence: 100% of MACHINE_OBSERVATION records have hour = 0
  # These are acoustic/satellite tags defaulting to midnight, not sightings
  n_total   <- nrow(occurrence_data)
  n_machine <- sum(occurrence_data$record_type == "MACHINE_OBSERVATION",
                   na.rm = TRUE)

  occ <- occurrence_data %>%
    filter(record_type != "MACHINE_OBSERVATION")

  if (n_machine > 0) {
    message("Excluded ", n_machine, "/", n_total,
            " MACHINE_OBSERVATION records (all recorded at hour 0 by default)")
  }

  # ── 2. Remove hour = 0 ────────────────────────────────────────────────────
  # Hour 0 in human observations also likely reflects missing time data,
  # not genuine midnight sightings
  n_zero <- sum(occ$hour == 0, na.rm = TRUE)
  occ    <- occ %>% filter(!is.na(hour), hour != 0)

  if (n_zero > 0) {
    message("Excluded ", n_zero,
            " records with hour = 0 (likely missing time, not genuine midnight sightings)")
  }

  # ── 3. Derive ecological plausibility window from the data ─────────────────
  # Central 80% of sightings defines the plausible hour range
  # No hardcoding — works for any organism
  hour_dist <- occ %>%
    count(hour) %>%
    arrange(hour) %>%
    mutate(cum_pct = cumsum(n) / sum(n))

  plausible_hours <- hour_dist %>%
    filter(cum_pct >= 0.10, cum_pct <= 0.90) %>%
    pull(hour)

  plaus_min <- min(plausible_hours)
  plaus_max <- max(plausible_hours)

  message("Ecological plausibility window (central 80% of sightings): ",
          sprintf("%02d:00", plaus_min), " - ", sprintf("%02d:00", plaus_max))

  # Cosine taper outside plausibility window — no hard cliff at boundary
  plausibility_score <- function(h) {
    dist <- pmax(0, pmax(plaus_min - h, h - plaus_max))
    ifelse(h >= plaus_min & h <= plaus_max,
           1.0,
           pmax(0.2, cos(dist / 3 * pi / 2)))
  }

  # ── 4. Get weather for top stations, averaged per date ─────────────────────
  org_stations <- top_stations %>%
    filter(organism == organism_name)

  if (nrow(org_stations) == 0) {
    stop("No top stations found for organism: ", organism_name)
  }

  weather_org <- weather_data %>%
    filter(ws_id %in% org_stations$ws_id) %>%
    group_by(date, month, dayofyear) %>%
    summarise(
      temp       = mean(temp,       na.rm = TRUE),
      rh         = mean(rh,         na.rm = TRUE),
      prcp       = mean(prcp,       na.rm = TRUE),
      rainy      = mean(rainy,      na.rm = TRUE),
      wind_speed = mean(wind_speed, na.rm = TRUE),
      .groups    = "drop"
    ) %>%
    # mean(na.rm = TRUE) on all-NA groups produces NaN — convert to NA
    mutate(across(where(is.numeric), ~ifelse(is.nan(.x), NA, .x)))

  # ── 5. Join occurrences to weather ────────────────────────────────────────
  occ_weather <- occ %>%
    select(date, hour, month, weekday, dayofyear) %>%
    inner_join(weather_org, by = "date")

  # ── 6. Self-calibrating weather weights via Cohen's d ─────────────────────
  # Cohen's d measures how much each weather variable differs between
  # sighting days and all days. Variables that differ more get more weight.
  cohen_d <- function(x_obs, x_all) {
    x_obs <- x_obs[!is.na(x_obs)]
    x_all <- x_all[!is.na(x_all)]
    if (length(x_obs) < 3 || length(x_all) < 3) return(0)
    s <- sd(x_all)
    if (is.na(s) || s == 0) return(0)
    abs(mean(x_obs) - mean(x_all)) / s
  }

  d_temp  <- cohen_d(occ_weather$temp,       weather_org$temp)
  d_rain  <- cohen_d(occ_weather$rainy,      weather_org$rainy)
  d_wind  <- cohen_d(occ_weather$wind_speed, weather_org$wind_speed)
  d_rh    <- cohen_d(occ_weather$rh,         weather_org$rh)
  total_d <- d_temp + d_rain + d_wind + d_rh + 1e-6

  ideal_temp  <- mean(occ_weather$temp,       na.rm = TRUE)
  ideal_rain  <- mean(occ_weather$rainy,      na.rm = TRUE)
  ideal_wind  <- mean(occ_weather$wind_speed, na.rm = TRUE)
  ideal_rh    <- mean(occ_weather$rh,         na.rm = TRUE)

  # Score each historical weather day by closeness to ideal conditions
  weather_scored <- weather_org %>%
    mutate(
      weather_score = 1 - (
        (d_temp / total_d) * pmin(abs(temp       - ideal_temp) / 10,  1) +
        (d_rain / total_d) * pmin(abs(rainy       - ideal_rain),       1) +
        (d_wind / total_d) * pmin(abs(wind_speed  - ideal_wind) / 10,  1) +
        (d_rh   / total_d) * pmin(abs(rh          - ideal_rh)   / 100, 1)
      ),
      weather_score = ifelse(is.na(weather_score), 0.5, weather_score)
    ) %>%
    select(date, month, weather_score)

  monthly_weather <- weather_scored %>%
    group_by(month) %>%
    summarise(avg_weather_score = mean(weather_score, na.rm = TRUE),
              .groups = "drop")

  # ── 7. Frequency scores ────────────────────────────────────────────────────
  month_scores <- occ %>%
    count(month, name = "n_month") %>%
    mutate(month_score = n_month / sum(n_month))

  hour_scores <- occ %>%
    count(hour, name = "n_hour") %>%
    mutate(hour_score = n_hour / sum(n_hour))

  # ── 8. Confidence: sightings per month x hour cell ────────────────────────
  month_hour_counts <- occ %>%
    count(month, hour, name = "n_cell")

  # ── 9. Build month x hour grid and score every combination ────────────────
  grid <- merge(data.frame(month = 1:12), data.frame(hour = 1:23))

  results <- grid %>%
    left_join(month_scores,      by = "month") %>%
    left_join(monthly_weather,   by = "month") %>%
    left_join(hour_scores,       by = "hour")  %>%
    left_join(month_hour_counts, by = c("month", "hour")) %>%
    replace_na(list(
      month_score       = 0,
      avg_weather_score = 0.5,
      hour_score        = 0,
      n_cell            = 0
    )) %>%
    mutate(
      # Base composite score
      composite_score = 0.40 * month_score +
                        0.40 * hour_score  +
                        0.20 * avg_weather_score,

      # Confidence: exponential saturation curve
      # 1 sighting = 0.18, 5 = 0.63, 10 = 0.86, 20+ approaches 0.98
      confidence   = 1 - exp(-n_cell / conf_k),

      # Ecological plausibility: data-derived window with cosine taper
      plausibility = plausibility_score(hour),

      # Final score: composite scaled by confidence-plausibility blend
      # Multiplicative so low confidence AND low plausibility compounds penalty
      # But low confidence WITH high plausibility still surfaces the result
      confidence_weight = 0.6 * confidence + 0.4 * plausibility,
      final_score       = composite_score * confidence_weight,

      # Labels
      month_name   = month.name[month],
      time_label   = sprintf("%02d:00", hour),
      period       = case_when(
        hour < 12 ~ "Morning",
        hour < 17 ~ "Afternoon",
        TRUE      ~ "Evening/Night"
      ),

      # Plain-English confidence interpretation
      confidence_tag = case_when(
        confidence >= 0.63 & plausibility >= 0.8 ~
          "High — strong data, ecologically expected",
        confidence >= 0.18 & plausibility >= 0.8 ~
          "Moderate — sparse data but ecologically plausible",
        confidence >= 0.63 & plausibility <  0.8 ~
          "Caution — data present but unusual time",
        TRUE ~
          "Low — sparse data and atypical time"
      )
    ) %>%
    arrange(desc(final_score))

  top_results <- results %>% slice_head(n = n)

  # ── 10. Print results ──────────────────────────────────────────────────────
  cat("\n")
  cat("  Best times to spot:", organism_name, "\n")
  cat(rep("-", 60), "\n", sep = "")
  cat(sprintf("  %-22s  %-6s  %-6s  %-6s  %s\n",
              "When", "Score", "Conf.", "Plaus.", "Tag"))
  cat(rep("-", 60), "\n", sep = "")

  for (i in seq_len(nrow(top_results))) {
    row <- top_results[i, ]
    cat(sprintf("  %d. %-10s %s  %.3f  %.2f    %.2f    %s\n",
                i,
                row$month_name,
                row$time_label,
                row$final_score,
                row$confidence,
                row$plausibility,
                row$confidence_tag))
  }

  cat("\n")
  cat("  Plausibility window:",
      sprintf("%02d:00", plaus_min), "-", sprintf("%02d:00", plaus_max), "\n")
  cat("  Cohen's d — temp:", round(d_temp, 3),
      "| rain:", round(d_rain, 3),
      "| wind:", round(d_wind, 3),
      "| rh:", round(d_rh, 3), "\n")
  cat("  Confidence k =", conf_k,
      "(1 sighting = 0.18, 5 = 0.63, 10 = 0.86, 20+ ~ 0.98)\n")

  invisible(list(
    recommendations     = top_results,
    full_grid           = results,
    plausibility_window = c(plaus_min, plaus_max),
    cohen_d = c(temp = d_temp, rain = d_rain, wind = d_wind, rh = d_rh)
  ))
}
```

------------------------------------------------------------------------

## Results

### Glowworms

```{r}
#| label: results-glowworms

results_glow <- predict_best_spotting_times(
  glowworms, weather, top_stations, "glowworms"
)
```

December dominates, which reflects the combined signal from all three
states. Notably, December 15:00 ranks first — initially surprising for a
bioluminescent organism. Investigation of the state-level data reveals
why:

```{r}
#| label: glowworm-state-hour

glowworms %>%
  filter(month == 12, hour != 0) %>%
  mutate(
    time_type = case_when(
      hour >= 20 | hour <= 4 ~ "Night",
      hour >= 17             ~ "Evening",
      TRUE                   ~ "Daytime"
    )
  ) %>%
  count(obs_state, time_type) %>%
  arrange(obs_state, time_type)
```

Tasmania accounts for the majority of December sightings, and 19 of
those are daytime — almost certainly organised cave and tunnel tours
where artificial darkness makes glowworms visible regardless of the time
outside. Queensland sightings, by contrast, cluster after 18:00 in line
with natural bioluminescence behaviour.

The confidence scoring system surfaced this result rather than
suppressing it. A simple frequency ranking would have ranked December
22:00 first and moved on. The high confidence score at 15:00 prompted
the investigation that revealed two genuinely different spotting
contexts within one organism dataset — cave tourism in Tasmania and wild
sightings in Queensland.

The plausibility window of 09:00–21:00 is broader than expected for a
nocturnal organism precisely because it is data-derived and both
contexts are present in the data.

### Gouldian Finch

```{r}
#| label: results-finch

results_finch <- predict_best_spotting_times(
  gouldian_finch, weather, top_stations, "gouldian_finch"
)
```

All five top recommendations are at 07:00 across the dry season months
(May–September). Confidence is 1.00 across the board — this is the most
data-rich dataset with a median of 8 sightings per month × hour cell.
The dawn chorus signal is unambiguous and the weather effect sizes
(Cohen's d for temp = 0.40, rh = 0.38) confirm that dry season
conditions are meaningfully associated with sightings.

### Manta Rays

```{r}
#| label: results-manta

results_manta <- predict_best_spotting_times(
  manta_rays, weather, top_stations, "manta_rays"
)
```

The function automatically excludes 675 machine observation records
before any analysis. After exclusion, June emerges as the peak month
with mid-morning hours (08:00–11:00) scoring highest — consistent with
divers and snorkellers operating during daylight in winter feeding
aggregations.

June 22:00 appears at rank 4 with a "Caution" flag (confidence 0.97 but
plausibility 0.50). This is the system working correctly: there are real
human sightings recorded at 22:00 in June, but the hour sits outside the
central 80% of the distribution. The recommendation surfaces with a
clear warning rather than being silently suppressed or silently
promoted.

The temperature Cohen's d of 0.58 is the highest of any organism —
weather is a genuine predictor for manta rays, reflecting their
sensitivity to water temperature during feeding season.

### Orchids

```{r}
#| label: results-orchids

results_orchid <- predict_best_spotting_times(
  orchids, weather, top_stations, "orchids"
)
```

September dominates with all five slots falling between 10:00 and 14:00
— Australian spring bloom, mid-morning when pollinators are most active.
Confidence is 1.00 throughout; orchids have a median of 20.5 sightings
per month × hour cell, making this the most statistically robust set of
recommendations. Wind and humidity Cohen's d values are both 0,
confirming that for orchids, season and time of day carry the prediction
entirely — weather conditions on the day matter very little.

------------------------------------------------------------------------

## What makes this approach different

Three things distinguish this function from simpler approaches.

**Confidence scoring.** A recommendation based on 1 sighting looks
identical to one based on 50 in a frequency table. The exponential
saturation curve `1 - exp(-n/k)` makes this visible. The curve shape is
borrowed from reliability engineering — confidence builds quickly at
first then plateaus, which is the right behaviour for count data. The k
parameter is tunable: a smaller k rewards sparse data more generously, a
larger k is stricter.

**Data-derived ecological plausibility.** Rather than hardcoding
activity windows per organism, the function calculates the central 80%
of the hour distribution and applies a cosine taper outside it. This
works for any organism without prior knowledge. The cosine taper means
there is no abrupt penalty at the boundary — an hour just outside the
window still receives partial credit.

**Multiplicative combination.** The three components — composite score,
confidence, and plausibility — combine multiplicatively rather than
additively. This means low confidence AND low plausibility compounds the
penalty. But a low-confidence recommendation that is ecologically
plausible still surfaces, labelled honestly as "Moderate — sparse data
but ecologically plausible." The plain-English confidence tag means the
output is immediately interpretable without reading the underlying
numbers.

The manta ray June 22:00 result and the glowworm December 15:00 result
are both examples of the system behaving correctly — one surfaced with a
caution flag, one surfaced with high confidence and prompted an
investigation that revealed a genuine biological and ecological pattern
in the data.
